---
title: "Asthma Brief V2"
author: "MikaylaBranz"
date: "9/7/2020"
output: 
  html_notebook:
    toc: yes
    toc_float: yes
    code_folding: "hide"
---

```{r setup, include=FALSE, echo = FALSE}
knitr::opts_chunk$set(echo = TRUE, include = TRUE, message =FALSE, warning = FALSE)
```
# Bring in Data
Set up data frame:

- Import data from multiple CSV files
- Combine into one data frame
- Trim to remove extraneous variables
```{r, import data, eval = FALSE}

#Open packages 
library(readr)
library(tidyverse)
library(ggplot2)

#Use readr to import csv files. 
asthma2014 <- read_csv("2014_Asthma.csv")
asthma2015 <- read_csv("2015_Asthma.csv")
asthma2016 <- read_csv("2016_Asthma.csv")
asthma2017 <- read_csv("2017_Asthma.csv")
asthma2018 <- read_csv("2018_Asthma.csv")
asthma2019 <- read_csv("2019_Asthma.csv")
asthma2020 <- read_csv("2020_Asthma_9-2.csv")

#combine datasets
#make list of datasets to combine
asthmaAll <- list (asthma2014,
                asthma2015,
                asthma2016,
                asthma2017,
                asthma2018,
                asthma2019,
                asthma2020)

#combine elements in the list
asthma <- 
  asthmaAll %>%
  bind_rows()

#select only variables that are relevant
asthma <- asthma %>% 
  select(Date, Zipcode,Age, Sex, Hospital,Race_flat,Ethnicity_Flat, HospitalName,HospitalZipCode,EssenceID, PIN, Time)
```
# Clean Dataset

- Separate date into month, day, and year
- Fix variable classes
- Recode race  
- Recode ethnicity
- Recode age
- Remove duplicates and case with impossible age (117)
- Remove cases from September 2020 (not a complete month)
```{r, eval = FALSE, echo= FALSE}
#Separate date into month, day, and year

asthma <- asthma %>%
  separate(col = Date, into = c("Month", "Day", "Year"), sep = "/", remove = FALSE)  

#Check and fix variable classes

class(asthma)
sapply(asthma,class)

asthma <- data.frame(asthma) 
asthma$Month <- as.numeric(asthma$Month)
asthma$Day <- as.numeric(asthma$Day)
asthma$Year <- as.numeric(asthma$Year)
asthma$Zipcode <- as.factor(asthma$Zipcode)
asthma$Age <- as.numeric(asthma$Age)
asthma$Sex <- as.factor(asthma$Sex)
asthma$Race_flat <- as.factor(asthma$Race_flat)
asthma$Ethnicity_Flat <- as.factor(asthma$Ethnicity_Flat)

#Recode Ethnicity
asthma <- asthma %>% 
  mutate(Ethnicity = recode(Ethnicity_Flat,
         ";Non-Hispanic;" = "NonHispanic", 
         ";NOT H;" = "NonHispanic", 
         ";Not H;" = "NonHispanic",
         ";Not Hispanic or Latino;" = "NonHispanic",
         
         ";Unknown;" = "UnknownOrNotReported",
         ";N;" = "UnknownOrNotReported", 
         ";U;" ="UnknownOrNotReported", 
         ";NR;" = "UnknownOrNotReported", 
         ";BL;" = "UnknownOrNotReported", 
         ";WH;" = "UnknownOrNotReported",
         
         ";Hispanic;" = "HispanicLatino",
         ";Hispa;" = "HispanicLatino", 
         ";H;" = "HispanicLatino",
         ";Hispanic or Latino;" = "HispanicLatino")) 


#Recode Race
#Race includes categories (BlackOrAfricanAmerican, White, Other, Unknown, AsianOrPacificIslander, AmericanIndianOrAlaskaNative, MixedRace)
#Racecondensed includes categories (Black, White, Other)
asthma <- asthma %>% 
  mutate(Race = recode(Race_flat,
                       ";Black;" = "BlackOrAfricanAmerican",
                       ";B;" = "BlackOrAfricanAmerican",
                       ";Black or African American;" = "BlackOrAfricanAmerican",
                       ";Black or African American;Black or African American;" = "BlackOrAfricanAmerican",
                       
                       ";White;" = "White",
                       ";W;" = "White",
                       
                       ";Other;" = "Other",
                       ";Ameri;" = "Other",
                       ";O;" = "Other",
                       ";Other Race;" = "Other",
                       ";Other Race;Other Race;" = "Other",
                       
                       ";Unknown;" = "Unknown",
                       ";NR;" = "Unknown",
                       ";U;" = "Unknown",
                       
                       ";Asian or Pacific Islander;" = "AsianOrPacificIslander",
                       ";Asian;" = "AsianOrPacificIslander",
                       ";A;" = "AsianOrPacificIslander",
                       ";Native hawaiian or Other Pacific Islander;" = "AsianOrPacificIslander",
                       
                       ";American Indian or Alaskan Native;" = "AmericanIndianOrAlaskaNative",
                       ";Nativ;" = "AmericanIndianOrAlaskaNative",
                       ";American Indian or Alaska Native;" = "AmericanIndianOrAlaskaNative",
                       
                       ";M;" = "MixedRace",
                       ";White;Black or African American;" = "MixedRace",
                       ";Black or African American;White;" = "MixedRace",
                       ";Other Race;White;" = "MixedRace")) %>% 
  mutate(RaceCondensed = recode(Race, 
                               "BlackOrAfricanAmerican" = "Black",
                               "White" = "White",
                               "Other" = "Other",
                               "Unknown" = "Other",
                               "AsianOrPacificIslander" = "Other",
                               "AmericanIndianOrAlaskaNative" = "Other",
                               "MixedRace" = "Other"))

#Recode age into categories
asthma <- asthma %>% 
  mutate(AgeGroup =  if_else(Age %in% 0:9, "0-9",
                     ifelse(Age %in% 10:19, "10-19",
                     ifelse(Age %in% 20:29, "20-29",
                     ifelse(Age %in% 30:39, "30-39",
                     ifelse(Age %in% 40:49, "40-49",
                     ifelse(Age %in% 50:59, "50-59",
                     ifelse(Age %in% 60:69, "60-69",
                     ifelse(Age %in% 70:79, "70-79",
                     ifelse(Age >= 80, "80+", "NA"))))))))))

#remove any duplicates in data
asthma <- distinct(asthma, .keep_all = FALSE)

#Remove case with impossible age (117)
asthma <- asthma %>% 
  filter(Age != "117")

#Remove cases from September 2020
asthma <- asthma %>% 
  filter(!(Year == "2020" & Month == 9))
```
# Proportions: add columns that show the populations by:

- Race (RacePop)
- Race for children 0-10 (Under10Pop)
- Age group (AgeGroupPop)
- Ethnicity (EthnicityPop)
```{r, eval=FALSE}
#Import race populations
RacePopulation <- read_csv("RacePopulation.csv")

#Check and fix class of RacePopulation dataframe and Race variable
class(RacePopulation)
RacePopulation<-data.frame(RacePopulation)

class(RacePopulation$Race)
RacePopulation$Race<-as.factor(RacePopulation$Race)


#check and fix race variable names to match between RacePopulation and asthma datasets
unique(RacePopulation$Race)
unique(asthma$Race)

RacePopulation$Race <- recode(RacePopulation$Race, "African American" = "Black", "Caucasian" = "White")

#add column to asthma that includes population by race
asthma <- asthma %>% 
  left_join(RacePopulation, by = c("RaceCondensed" = "Race")) %>% 
    rename("RacePop" = "Population")

#Code for children 0-10 by race
ChildrenRacePopulation <- read_csv("Under10RacePopulation.csv")

#Check and fix class of RacePopulation dataframe and Race variable
class(ChildrenRacePopulation)
ChildrenRacePopulation<-data.frame(ChildrenRacePopulation)

class(ChildrenRacePopulation$Race)
ChildrenRacePopulation$Race<-as.factor(ChildrenRacePopulation$Race)


#check and fix variable names to match
unique(ChildrenRacePopulation$Race)
unique(asthma$RaceCondensed)

ChildrenRacePopulation$Race <- recode(ChildrenRacePopulation$Race, "BlackOrAfricanAmerican" = "Black")

asthma <- asthma %>% 
  left_join(ChildrenRacePopulation, by = c("RaceCondensed" = "Race")) %>% 
    rename("Under10Pop" = "Under10Pop")

#Import Age Group populations
AgeGroupPopulation <- read_csv("AgeGroupPopulation.csv")

#Check and fix class of RacePopulation dataframe and Race variable
class(AgeGroupPopulation)
AgeGroupPopulation<-data.frame(AgeGroupPopulation)

class(AgeGroupPopulation$AgeGroup)
AgeGroupPopulation$AgeGroup<-as.factor(AgeGroupPopulation$AgeGroup)


#check and fix variable names to match
unique(AgeGroupPopulation$AgeGroup)
unique(asthma$AgeGroup)

class(AgeGroupPopulation$AgeGroup)

asthma <- asthma %>% 
  left_join(AgeGroupPopulation, by = "AgeGroup") %>% 
    rename("AgeGroupPop" = "Population")

#Ethnicity Rate:

#Import Age Group populations
EthnicityPopulation <- read_csv("EthnicityPopulation.csv")

#Check and fix class of RacePopulation dataframe and Race variable
class(EthnicityPopulation)
EthnicityPopulation<-data.frame(EthnicityPopulation)

class(EthnicityPopulation$Ethnicity)
EthnicityPopulation$Ethnicity<-as.factor(EthnicityPopulation$Ethnicity)


#check and fix variable names to match
unique(EthnicityPopulation$Ethnicity)
unique(asthma$EthnicityCondensed)

class(EthnicityPopulation$Ethnicity)

asthma <- asthma %>% 
  mutate(EthnicityCondensed = recode(Ethnicity,
         "UnknownOrNotReported" = "NonHispanic")) 

EthnicityPopulation$Ethnicity <- recode(EthnicityPopulation$Ethnicity, "Non-Hispanic" = "NonHispanic", "Hispanic" = "HispanicLatino")
# wrong number

asthma <- asthma %>% 
  left_join(EthnicityPopulation, by = c("EthnicityCondensed" = "Ethnicity")) %>% 
    rename("EthnicityPop" = "Population")
```

# Graph Asthma ER visits by Race

## Graph Race by Case Count

```{r}
#change order of race in RaceCondensed
#make a list with new order of levels
 newlevel <- c("Black", "White","Other")
 #change the order of the levels in race
 asthma$RaceCondensed <- factor(asthma$RaceCondensed, levels = newlevel)

#Graph the total number of cases by race, over time. Use categories "Black, White, and Other" Only include Jan-August so that it is comparable to 2020
asthma %>% 
  filter(Month < 9) %>% 
  group_by(RaceCondensed, Year) %>% 
  summarise(total = n()) %>% 
  ggplot() +
   geom_line(aes(x = Year, y = total, color = RaceCondensed), size = 1) +
  labs(y="Cases", x = "Year", color = "Race", caption = "*2020 data represents Jan-Aug") +
   theme(axis.text.x=element_text(angle=45,  hjust = 1)) +
    scale_x_continuous(breaks = seq(2014,2020, by =1))+
   theme(panel.grid.minor.x=element_blank())+
  ggtitle("Asthma-Related ER Visits by Race (Jan-Aug), 2014-2020")

#Graph the total number of cases by race, over time. Use all race categories. Only include Jan-August so that it is comparable to 2020
asthma %>% 
  filter(Month < 9) %>% 
  group_by(Race, Year) %>% 
  summarise(total = n()) %>% 
  ggplot() +
   geom_line(aes(x = Year, y = total, color = Race), size = 1) +
  labs(y="Cases", x = "Year", color = "Race", caption = "*2020 data represents Jan-Aug") +
   theme(axis.text.x=element_text(angle=45,  hjust = 1)) +
    scale_x_continuous(breaks = seq(2014,2020, by =1))+
   theme(panel.grid.minor.x=element_blank())+
  ggtitle("Asthma-Related ER Visits by Race (Jan-Aug), 2014-2020")

#Graph the total number of cases by race, over time. Use categories "Black, White, and Other." Include all months even though data for 2020 is not complete.
asthma %>% 
  group_by(RaceCondensed, Year) %>% 
  summarise(total = n()) %>% 
  ggplot() +
   geom_line(aes(x = Year, y = total, color = RaceCondensed), size = 1) +
  labs(y="Cases", x = "Year", color = "Race", caption = "*2020 data represents Jan-Aug") +
   theme(axis.text.x=element_text(angle=45,  hjust = 1)) +
    scale_x_continuous(breaks = seq(2014,2020, by =1))+
   theme(panel.grid.minor.x=element_blank())+
  ggtitle("Asthma-Related ER Visits by Race, 2014-2020")

#Graph the total number of cases by race, over time. Use all race categories. Include all months even though data for 2020 is not complete.
asthma %>% 
  group_by(Race, Year) %>% 
  summarise(total = n()) %>% 
  ggplot() +
   geom_line(aes(x = Year, y = total, color = Race), size = 1) +
  labs(y="Cases", x = "Year", color = "Race", caption = "*2020 data represents Jan-Aug") +
   theme(axis.text.x=element_text(angle=45,  hjust = 1)) +
    scale_x_continuous(breaks = seq(2014,2020, by =1))+
   theme(panel.grid.minor.x=element_blank())+
  ggtitle("Asthma-Related ER Visits by Race, 2014-2020")
```
## Graph Race by Rate
```{r}
#Graph asthma rates by race in 2019
asthma %>% 
  filter(Year == 2019) %>% 
  group_by(RaceCondensed) %>% 
  summarise(total = (10000* n()/(12*mean(RacePop)))) %>%
  ggplot() +
  geom_col(aes(x = RaceCondensed, y = total, fill = RaceCondensed)) +
  labs(y="Cases", x = "Race") +
   theme(axis.text.x=element_text(angle=45,  hjust = 1)) +
   theme(panel.grid.minor.x=element_blank())+
   scale_fill_manual(values = c("darkorchid4","turquoise4", "#02075d")) +
    scale_y_continuous(expand = c(0,0), limits = c(0,11)) +
  ggtitle("Asthma ER Visit Cases per 10,000 people per month, by Race, 2019")

#Graph asthma rates by race in 2020
asthma %>% 
  filter(Year == 2020) %>% 
  group_by(RaceCondensed) %>% 
  summarise(total = (10000* n()/(8*mean(RacePop)))) %>%
  ggplot() +
  geom_col(aes(x = RaceCondensed, y = total, fill = RaceCondensed)) +
  labs(y="Cases", x = "Race", Caption = "*Jan-Aug, 2020") +
   theme(axis.text.x=element_text(angle=45,  hjust = 1)) +
   theme(panel.grid.minor.x=element_blank())+
  scale_fill_manual(values = c("darkorchid4","turquoise4", "#02075d")) +
  scale_y_continuous(expand = c(0,0), limits = c(0,11)) +
  ggtitle("Asthma ER Visit Cases per 10,000 people per month, by Race, 2020")

#Graph asthma rates by race in 2019 and 2020 combined 

Asthma2019RaceRate <- data.frame(asthma %>% 
  filter(Year == 2019) %>% 
  group_by(RaceCondensed, Year) %>% 
  summarise(total = (10000* n()/(12*mean(RacePop)))))

Asthma2020RaceRate <- data.frame(asthma %>% 
  filter(Year == 2020) %>% 
  group_by(RaceCondensed, Year) %>% 
  summarise(total = (10000* n()/(8*mean(RacePop)))))

#make list of datasets to combine
asthmaRaceAll <- list (Asthma2019RaceRate, Asthma2020RaceRate)

#Combine datasets
AsthmaRaceRate <- 
  asthmaRaceAll %>%
  bind_rows()

AsthmaRaceRate$Year <-as.factor(AsthmaRaceRate$Year)
#Now graph them together
AsthmaRaceRate %>% 
  ggplot() +
  geom_col(aes(x = Year, y=total, fill = RaceCondensed), position = "dodge") +
  labs(y="Cases per 10,000 people per month", x = "Year", fill = "Race") +
   theme(axis.text.x=element_text(angle=45,  hjust = 1)) +
   theme(panel.grid.minor.x=element_blank())+
   scale_fill_manual(values = c("darkorchid4","turquoise4", "#02075d")) +
    scale_y_continuous(expand = c(0,0), limits = c(0,11)) +
  ggtitle("Asthma ER Visit Cases per 10,000 people per month, by Race, 2019 and 2020")

#Graph by race and rate, 2014-2019
asthma %>% 
  filter(Year < 2020) %>% 
  group_by(RaceCondensed, Year) %>% 
  summarise(Total = n(), Pop = mean(RacePop), Rate = 1000*Total/Pop) %>% 
  ggplot() +
  geom_point(aes(x = Year, y = Rate, color = RaceCondensed), size = 3, shape = 4)+
   geom_line(aes(x = Year, y = Rate, color = RaceCondensed), size = 1) +
  labs(y="Cases per 1,000 People", x = "Year", color = "Race") +
   theme(axis.text.x=element_text(angle=45,  hjust = 1)) +
    scale_x_continuous(breaks = seq(2014,2020, by =1))+
   theme(panel.grid.minor.x=element_blank())+
  scale_color_manual(values = c("darkorchid4","turquoise4", "red"))+
  ggtitle("Asthma-Related ER Visits by Race, 2014-2019")
```
## Graph by Race and counts over time to check for drop off with COVID
```{r}
#Check on asthma ER Visits in 2020
asthma %>% 
  filter(Year == 2020) %>% 
  group_by(RaceCondensed, Month) %>% 
  summarise(total = n()) %>% 
  ggplot() +
   geom_line(aes(x = Month, y = total, color = RaceCondensed), size = 1) +
  labs(y="Cases", x = "Month", color = "Race") +
   theme(axis.text.x=element_text(angle=45,  hjust = 1)) +
    scale_x_continuous(breaks = seq(0,8, by =1))+
   theme(panel.grid.minor.x=element_blank())+
  ggtitle("Asthma ER Visit Cases by Month, 2020")

#Check on asthma ER Visits in 2020 for children
asthma %>% 
  filter(Year == 2020 & AgeGroup == "0-9") %>% 
  group_by(RaceCondensed, Month) %>% 
  summarise(total = n()) %>% 
  ggplot() +
   geom_line(aes(x = Month, y = total, color = RaceCondensed), size = 1) +
  labs(y="Cases", x = "Month", color = "Race") +
   theme(axis.text.x=element_text(angle=45,  hjust = 1)) +
    scale_x_continuous(breaks = seq(0,8, by =1))+
   theme(panel.grid.minor.x=element_blank())+
  ggtitle("0-9 year olds Asthma ER Visit Cases by Month, 2020")

#Check on asthma ER Visits in 2019 for comparison
asthma %>% 
  filter(Year == 2019) %>% 
  group_by(RaceCondensed, Month) %>% 
  summarise(total = n()) %>% 
  ggplot() +
   geom_line(aes(x = Month, y = total, color = RaceCondensed), size = 1) +
  labs(y="Cases", x = "Month", color = "Race") +
   theme(axis.text.x=element_text(angle=45,  hjust = 1)) +
    scale_x_continuous(breaks = seq(0,12, by =1))+
   theme(panel.grid.minor.x=element_blank())+
  ggtitle("Asthma ER Visit Cases by Month, 2019")

#Check on asthma ER Visits in 2019 of children 0-9 for comparison
asthma %>% 
  filter(Year == 2019 & AgeGroup == "0-9") %>% 
  group_by(RaceCondensed, Month) %>% 
  summarise(total = n()) %>% 
  ggplot() +
   geom_line(aes(x = Month, y = total, color = RaceCondensed), size = 1) +
  labs(y="Cases", x = "Month", color = "Race") +
   theme(axis.text.x=element_text(angle=45,  hjust = 1)) +
    scale_x_continuous(breaks = seq(0,12, by =1))+
   theme(panel.grid.minor.x=element_blank())+
  ggtitle("0-9 Year olds Asthma ER Visit Cases by Month, 2019")

#Put both years together

#Change Month to factor variable
asthma$MonthSpelled <- as.factor(asthma$Month)

asthma$MonthSpelled <-recode(asthma$MonthSpelled, "1" = "Jan", "2" = "Feb", "3" = "Mar", "4"= "Apr", "5" = "May", "6" = "Jun", "7" = "Jul", "8" = "Aug", "9" = "Sep", "10" = "Oct", "11" = "Nov", "12" = "Dec")

levels
#combine month and year
class(asthma$Year)
asthma$Year <- as.factor(asthma$Year)

asthma <- asthma %>%  
  unite(col= "MonthYear",
         MonthSpelled, Year, 
         sep = " ", 
         remove = FALSE)

#Make MonthYear a factor variable
asthma$MonthYear<- as.factor(asthma$MonthYear)

#check levels
levels(asthma$MonthYear)
#make a list with new order of levels
 MonthYearNewLevel <- c("Jan 2019", "Feb 2019", "Mar 2019", "Apr 2019", "May 2019", "Jun 2019", "Jul 2019", "Aug 2019", "Sep 2019", "Oct 2019", "Nov 2019", "Dec 2019", "Jan 2020", "Feb 2020", "Mar 2020", "Apr 2020", "May 2020", "Jun 2020", "Jul 2020", "Aug 2020")

#change the order of the levels in race
asthma$MonthYear <- factor(asthma$MonthYear, levels = MonthYearNewLevel)

#Graph asthma ER cases from 2019-2020
asthma %>% 
  filter(Year == 2019 | Year == 2020) %>% 
  group_by(MonthYear, RaceCondensed) %>% 
  summarise(Cases = n())%>% 
  ggplot() +
   geom_point(aes(x = MonthYear, y = Cases, color = RaceCondensed)) +
  geom_line(aes(x = MonthYear, y = Cases, color = RaceCondensed, group =RaceCondensed))+
  labs(y="Cases", x = "Month", color = "Race") +
   theme(axis.text.x=element_text(angle=45,  hjust = 1)) +
  scale_color_manual(values = c("darkorchid4","turquoise4", "#02075d")) +
  ggtitle("Asthma ER Visit Cases by Month, All Ages, 2019-2020")+
  geom_vline(xintercept = 15.5, linetype = "dashed", color = "red", size = 1.5)+
  geom_text(mapping = aes (x= 16.8, y =160, label = "COVID"), color = "red", size = 3.2)

asthma %>% 
  filter((Year == 2019 | Year == 2020) & AgeGroup == "0-9") %>% 
  group_by(MonthYear, RaceCondensed) %>% 
  summarise(Cases = n())%>% 
  ggplot() +
   geom_point(aes(x = MonthYear, y = Cases, color = RaceCondensed)) +
  geom_line(aes(x = MonthYear, y = Cases, color = RaceCondensed, group =RaceCondensed))+
  labs(y="Cases", x = "Month", color = "Race") +
   theme(axis.text.x=element_text(angle=45,  hjust = 1)) +
  scale_color_manual(values = c("darkorchid4","turquoise4", "#02075d")) +
  ggtitle("0-9 Year Olds Asthma ER Visit Cases by Month, All Ages, 2019-2020")+
  geom_vline(xintercept = 15.5, linetype = "dashed", color = "red", size = 1.5)+
  geom_text(mapping = aes (x= 16.8, y =60, label = "COVID"), color = "red", size = 3.2)
```

## Graph by race and rate over time to check for drop off with COVID
```{r}

#Graph asthma ER rates by race from 2019-2020
asthma %>% 
  filter(Year == 2019 | Year == 2020) %>% 
  group_by(MonthYear, RaceCondensed) %>% 
  summarise(Total = n(), Pop = mean(RacePop), Rate = 1000*Total/Pop) %>% 
  ggplot() +
   geom_point(aes(x = MonthYear, y = Rate, color = RaceCondensed)) +
  geom_line(aes(x = MonthYear, y = Rate, color = RaceCondensed, group =RaceCondensed))+
  labs(y="Cases per 1,000 people", x = "Month", color = "Race") +
   theme(axis.text.x=element_text(angle=45,  hjust = 1)) +
  scale_color_manual(values = c("darkorchid4","turquoise4", "#02075d")) +
  ggtitle("Asthma ER Visit Cases by Month, All Ages, 2019-2020")+
  geom_vline(xintercept = 15.5, linetype = "dashed", color = "red", size = 1.5)+
  geom_text(mapping = aes (x= 16.8, y =1.2, label = "COVID"), color = "red", size = 3.2)

#Graph asthma ER rates by race from 2019-2020 for children <10
# make NaN 0 so that 0 reported cases appears as 0
AsthmaRates<- asthma %>% 
  filter((Year == 2019 | Year == 2020)& Age <10)%>% 
  group_by(MonthYear, RaceCondensed, .drop = FALSE) %>% 
  summarise(Total = n(), Pop = mean(Under10Pop), Rate = 1000*Total/Pop) 
  
AsthmaRates$Rate[is.nan(AsthmaRates$Rate)]<-0

AsthmaRates%>% 
  ggplot() +
   geom_point(aes(x = MonthYear, y = Rate, color = RaceCondensed)) +
  geom_line(aes(x = MonthYear, y = Rate, color = RaceCondensed, group =RaceCondensed))+
  labs(y="Cases per 1,000 people", x = "Month", color = "Race") +
   theme(axis.text.x=element_text(angle=45,  hjust = 1)) +
  scale_color_manual(values = c("darkorchid4","turquoise4", "#02075d")) +
  ggtitle("Asthma ER Visit Cases by Month, Ages 0-9, 2019-2020")+
  geom_vline(xintercept = 15.5, linetype = "dashed", color = "red", size = 1.5)+
  geom_text(mapping = aes (x= 16.8, y =1.2, label = "COVID"), color = "red", size = 3.2)

```



```{r}
#Check the total number of cases by ethnicity, over time. Only include Jan-August so that it is comparable to 2020
asthma %>% 
  filter(Month < 9) %>% 
  group_by(Ethnicity, Year) %>% 
  summarise(total = n()) 
#There aren't really enough Latinx cases to be able to look at this as a proportion? 

#Graph the total number of cases by zipcode in 2019
asthma %>% 
  filter(Year==2019) %>% 
  group_by(Zipcode) %>% 
summarise(total = n()) %>% 
  ggplot() +
   geom_col(aes(x = reorder(Zipcode, -total), y= total, dodge = TRUE), fill= "Maroon") +
  theme(axis.text.x=element_text(angle=45, hjust=1), legend.position = "none") +
  scale_y_continuous(expand = c(0,0),
                     limits = c(0,300)) +
  labs(y="Cases", x = "Zip Code")+
  ggtitle("Asthma ER Visit Cases by Zip Code, Jan.- Dec. 2019")

#Graph the total number of cases by zipcode in 2020
asthma %>% 
  filter(Year==2020 | Year==2019) %>% 
  group_by(Zipcode, Year) %>% 
  summarise(total = n()) %>% 
summarise(total = n()) %>% 
  ggplot() +
   geom_col(aes(x = reorder(Zipcode, -total), y= total, dodge = TRUE), fill = "Maroon") +
  theme(axis.text.x=element_text(angle=45,hjust=1,), legend.position = "none") +
    scale_y_continuous(expand = c(0,0), limits = c(0,150)) +
  labs(y="Cases", x = "Zip Code")+
  ggtitle("Asthma ER Visit Cases by Zip Code, Jan.-Aug. 2020")


 
 #Now figure out how to do it as a proportion of the population.


```


## Graph by age group
```{r}

#Explore data by age group



#make AgeGroup factor variable
asthma$AgeGroup<-as.factor(asthma$AgeGroup)

asthma <- asthma %>% 
  mutate(AgeGroupPractical =  
                     if_else(Age %in% 0:5, "0-5",
                     ifelse(Age %in% 6:10, "6-10",
                     ifelse(Age %in% 11:14, "11-14",
                     ifelse(Age %in% 15:18, "15-18",
                     ifelse(Age %in% 19:26, "19-26",
                     ifelse(Age %in% 27:39, "27-39",
                     ifelse(Age %in% 40:59, "40-59",
                     ifelse(Age %in% 60:79, "60-79",
                     ifelse(Age >= 80, "80+", "NA"))))))))))

#make AgeGroup factor variable
asthma$AgeGroupPractical<-as.factor(asthma$AgeGroupPractical)

levels(asthma$AgeGroup)

#Graph the total number of cases by age group in 2019
asthma %>% 
  filter(Year==2019) %>% 
  group_by(AgeGroup) %>% 
summarise(total = n()) %>% 
  ggplot() +
   geom_col(aes(x = AgeGroup, y= total, dodge = TRUE), fill = "Navy") +
  theme(axis.text.x=element_text(angle=45,hjust=1,), legend.position = "none") +
  labs(y="Cases", x = "Age Group")+
  scale_y_continuous(expand = c(0,0), limits = c(0,450)) +
  ggtitle("Asthma ER Visit Cases by Age Group, Jan.-Dec. 2019")

#Graph the total number of cases by age group in 2020
asthma %>% 
  filter(Year==2020) %>% 
  group_by(AgeGroup) %>% 
summarise(total = n()) %>% 
  ggplot() +
   geom_col(aes(x = AgeGroup, y= total, dodge = TRUE), fill = "Navy") +
  theme(axis.text.x=element_text(angle=45,hjust=1,), legend.position = "none") +
  labs(y="Cases", x = "Age Group")+
 scale_y_continuous(expand = c(0,0), limits=c(0,200)) +
  ggtitle("Asthma ER Visit Cases by Age Group, Jan.-Aug 2020")

#Graph the total number of cases by practical age group in 2019
asthma %>% 
  filter(Year==2019) %>% 
  group_by(AgeGroupPractical) %>% 
summarise(total = n()) %>% 
  ggplot() +
   geom_col(aes(x = AgeGroupPractical, y= total, dodge = TRUE), fill = "Navy") +
  theme(axis.text.x=element_text(angle=45,hjust=1,), legend.position = "none") +
  labs(y="Cases", x = "Age Group")+
  scale_y_continuous(expand = c(0,0), limits = c(0,500)) +
  ggtitle("Asthma ER Visit Cases by Age Group, Jan.-Dec. 2019")

#Graph the total number of cases by age group in 2020
asthma %>% 
  filter(Year==2020) %>% 
  group_by(AgeGroupPractical) %>% 
summarise(total = n()) %>% 
  ggplot() +
   geom_col(aes(x = AgeGroupPractical, y= total, dodge = TRUE), fill = "Navy") +
  theme(axis.text.x=element_text(angle=45,hjust=1,), legend.position = "none") +
  labs(y="Cases", x = "Age Group")+
 scale_y_continuous(expand = c(0,0), limits=c(0,210)) +
  ggtitle("Asthma ER Visit Cases by Age Group, Jan.-Aug 2020")
       

```
```{r}
#Now to make proportions



#Age Group Proportions:

Ethnicity proportions

asthma %>% 
  filter(Month < 9) %>% 
  group_by(EthnicityCondensed, Year) %>% 
  summarise(total = (10000* n()/(8*mean(EthnicityPop)))) %>%
  ggplot() +
   geom_line(aes(x = Year, y = total, color = EthnicityCondensed), size = 1) +
  labs(y="Cases", x = "Year", color = "Ethnicity") +
   theme(axis.text.x=element_text(angle=45,  hjust = 1)) +
    scale_x_continuous(breaks = seq(2014,2020, by =1))+
   theme(panel.grid.minor.x=element_blank())+
  ggtitle("Asthma ER Visit Cases per 10,000 people per month, by Ethnicity, 2014-2020")

# NOTE: Sizes are probably too small for hispanic/latino to really be reasonable. Also, there are a LOT of unknowns, which got lumped into "not hispanic or latino" but are difficult to make meaningful conclusions from. I would NOT trust the analysis of ethnicity. 


```
```{r}
#Graph asthma rates by race over time
asthma %>% 
  filter(Month < 9) %>% 
  group_by(RaceCondensed, Year) %>% 
  summarise(total = (10000* n()/(8*mean(RacePop)))) %>%
  ggplot() +
   geom_line(aes(x = Year, y = total, color = RaceCondensed), size = 1) +
  labs(y="Cases", x = "Year", color = "Race") +
   theme(axis.text.x=element_text(angle=45,  hjust = 1)) +
    scale_x_continuous(breaks = seq(2014,2020, by =1))+
   theme(panel.grid.minor.x=element_blank())+
  ggtitle("Asthma ER Visit Cases per 10,000 people per month, by Race, 2014-2020")

#Graph asthma rates by race in 2020

asthma %>% 
  filter(Year == 2020) %>% 
  group_by(RaceCondensed) %>% 
  summarise(total = (10000* n()/(8*mean(RacePop)))) %>%
  ggplot() +
  geom_col(aes(x = RaceCondensed, y = total, fill = RaceCondensed)) +
  labs(y="Cases", x = "Race") +
   theme(axis.text.x=element_text(angle=45,  hjust = 1)) +
   theme(panel.grid.minor.x=element_blank())+
  scale_fill_manual(values = c("darkorchid4","turquoise4", "#02075d")) +
  ggtitle("Asthma ER Visit Cases per 10,000 people per month, by Race, 2020")

asthma %>% 
  filter(Year == 2019) %>% 
  group_by(RaceCondensed) %>% 
  summarise(total = (10000* n()/(12*mean(RacePop)))) %>%
  ggplot() +
  geom_col(aes(x = RaceCondensed, y = total, fill = RaceCondensed)) +
  labs(y="Cases", x = "Race") +
   theme(axis.text.x=element_text(angle=45,  hjust = 1)) +
   theme(panel.grid.minor.x=element_blank())+
   scale_fill_manual(values = c("darkorchid4","turquoise4", "#02075d")) +
  ggtitle("Asthma ER Visit Cases per 10,000 people per month, by Race, 2019")
```

```{r}
#Import age group population estimates
AgeGroupPopulation <- read_csv("AgeGroupPopulation.csv")

EthnicityPopulation <- read_csv("EthnicityPopulation.csv")
SexPopulation <- read_csv("SexPopulation.csv")
ZipcodePopulation <- read_csv("ZipcodePopulation.csv")

#Check and fix class

sapply(AgeGroupPopulation, class)
sapply(ZipcodePopulation, class)
sapply(RacePopulation, class)

left_join(AgeGroupPopulation, by = "AgeGroup") %>% 
   rename("AgePop" = "Population") %>%
```


















# Extra code
```{r}
RacePopulation$Race <- recode(RacePopulation$Race, "African American" = "Black", "Caucasian" = "White")

asthma <- asthma %>% 
  left_join(RacePopulation, by = c("RaceCondensed" = "Race")) %>% 
    rename("RacePop" = "Population")

#change order of race
#check levels
levels(asthma$RaceCondensed)
#make a list with new order of levels
 newlevel <- c("Black", "White","Other")
 #change the order of the levels in race
 asthma$RaceCondensed <- factor(asthma$RaceCondensed, levels = newlevel)
#Code for rate
RacePopulation <- read_csv("RacePopulation.csv")

#Check and fix class of RacePopulation dataframe and Race variable
class(RacePopulation)
RacePopulation<-data.frame(RacePopulation)

class(RacePopulation$Race)
RacePopulation$Race<-as.factor(RacePopulation$Race)


#check and fix variable names to match
unique(RacePopulation$Race)
unique(asthma$Race)
#Try for kids 0-9

#change order of race
#check levels
levels(asthma$RaceCondensed)
#make a list with new order of levels
 newlevel <- c("Black", "White","Other")
 #change the order of the levels in race
 asthma$RaceCondensed <- factor(asthma$RaceCondensed, levels = newlevel)
```
 
 